openTSNE: a modular \proglang{Python} library for t-SNE dimensionality reduction and embedding

### 2021-04-26 paper/main.tex
Remove pdf from repo, minor edits

The ever-growing volumes of high-dimensional data sets in machine learning calls
for efficient dimensionality reduction techniques and their implementations,
which can produce informative data visualizations. Popular approaches include
principal component analysis, multidimensional scaling, t-distributed stochastic
neighbor embedding (t-SNE)~\citep{maaten2008visualizing}, and uniform manifold
approximation and projections (UMAP)~\citep{2018arXivUMAP}. Among these, t-SNE
has received much attention as it can address high volumes of data and reveal
the underlying data structure. For instance, t-SNE is widely used in the
bioinformatics community in areas such as single-cell
transcriptomics~\citep{macosko2015highly,cao2019single,tasic2018shared}, human
genetics~\citep{hirata2019genetic}, metagenomic
assembly~\citep{beaulaurier2018metagenomic}, the spatial organization of
microbial communities~\citep{sheth2019spatial}, and
metabolomics~\citep{tkachev2019differences}. Reports on single-cell gene
expression data, our running example, often start with an overview of the cell
landscape, where t-SNE is used to embed high-dimensional expression profiles
into a two-dimensional space. Figs.~\ref{fig:macosko}.a and \ref{fig:macosko}.b
show two such embeddings.

@@

Despite its utility, t-SNE has often been criticized for its limited
scalability, lack of global organization -- t-SNE identifies well-defined
clusters that may be arbitrarily scattered throughout the low-dimensional space
-- and the absence of theoretically-founded methods to map new data into
existing embeddings~\citep{ding2018interpretable,becht2019dimensionality}. Most
of these shortcomings have recently been addressed. \citet{linderman2019fast}
developed FIt-SNE, an efficient approximation scheme which massively improves
the scalability of t-SNE, achieving linear time complexity in the number of
samples. \citet{kobak2019art} proposed several techniques to improve global
cluster coherence, including estimating similarities with a mixture of Gaussian
kernels. In our previous work, we introduced a principled approach for embedding
new samples into existing visualizations~\citep{policar2019embedding}.

@@

\subsection{t-distributed stochastic neighbor embedding} \label{sec:meth.tsne}

t-distributed stochastic neighbor embedding (t-SNE) is a non-linear
dimensionality reduction method that to aims to find a low-dimensional embedding
where neighborhoods are preserved. More formally, given a multi-dimensional data set
$\mathbf{X} = \left \{ \mathbf{x}_1, \mathbf{x}_2, \dots, \mathbf{x}_N \right \}
\in \mathbb{R}^D$ where $N$ is the number of data points in the data set, t-SNE
aims to find a low dimensional embedding $\mathbf{Y} = \left \{ \mathbf{y}_1,
\mathbf{y}_2, \dots, \mathbf{y}_N \right\} \in \mathbb{R}^d$ where $d \ll D$,
such that if points $\mathbf{x}_i$ and $\mathbf{x}_j$ are close in the
high-dimensional space, their corresponding embeddings $\mathbf{y}_i$ and
$\mathbf{y}_j$ are also close. Since t-SNE is primarily used as a visualization
tool, $d$ is typically set to two. The similarity between two data points in the
high-dimensional space is defined using the Gaussian kernel
\begin{equation}
p_{j \mid i} = \frac{\exp \left ( -\frac{1}{2} \mathcal{D}(\mathbf{x}_i, \mathbf{x}_j ) / \sigma_i^2 \right )}
{\sum_{k \neq i } \exp \left ( -\frac{1}{2} \mathcal{D}(\mathbf{x}_i, \mathbf{x}_k ) / \sigma_i^2 \right )}, \quad p_{i \mid i} = 0 \notag
\label{eq:gaussian_kernel}
\end{equation}
where $\mathcal{D}$ is some distance measure. The values $p_{j \mid i}$
are then symmetrized to
\begin{equation}
p_{ij} = \frac{p_{j \mid i} + p_{i \mid j}}{2N}.
\label{eq:symmetrize}
\end{equation}

The bandwidth $\sigma_i$ of each Gaussian kernel is selected such that the perplexity of the distribution matches a user-specified parameter value
\begin{equation}
\text{Perplexity} = 2^{H(P_i)} \notag
\end{equation}
where $H(P_i)$ is the Shannon entropy of $P_i$,
\begin{equation}
H(P_i) = -\sum_i p_{j \mid i} \log_2 (p_{j \mid i}). \notag
\end{equation}
This enables t-SNE to adapt to the varying density of the data in the
multi-dimensional space. The perplexity can be interpreted as as the continuous
analogue to the number of nearest neighbors to which the distances will be
preserved.

@@

Optimization is performed using batch gradient descent using the delta-bar-delta
update rule~\citep{jacobs1988increased}. The t-SNE optimization proceedure
consists of two phases: in the first \textit{early exaggeration} phase, the
attractive forces between data points are increased by some factor $\rho$,
typically set to $12$, so that points in the embedding can more easily move
throughout the space and settle near their respective neighbors. In the second
phase of the optimization, the attractive forces are then reverted to their
original values with $\rho=1$.

@@

A direct evaluation of t-SNE gradients requires $\mathcal{O}(N^2)$ operations,
which makes its application impractical to any reasonably-sized data set and
highlights the need for the development of efficient approximation schemes.
\citet{van2014accelerating} observed that the t-SNE gradient can be cast as an
N-body problem where data points represent particles that attract and repel each
other. The gradient from Eqn.~(\ref{eq:tsne_gradient}) can be rewritten as
\begin{equation}
\frac{\partial C}{\partial \mathbf{y}_i} = 4 \Bigg [
\underbrace{\sum_{j \neq i} p_{ij} q_{ij} Z \left ( \mathbf{y}_i - \mathbf{y}_j \right )}_{\text{attractive forces}}  -
\underbrace{\sum_{j \neq i} q_{ij}^2 Z \left ( \mathbf{y}_i - \mathbf{y}_j \right )}_{\text{repulsive forces}}
\Bigg ], \label{eq:grad_attr_rep} \notag
\end{equation}
where $Z = \sum_{k \neq l}\left ( 1 + || \mathbf{y}_k - \mathbf{y}_l ||^2 \right
)^{-1}$. We can view this equation as a particle simulation, where the
two terms represent the attractive and repulsive forces between individual
particles. Each term lends itself to efficient approximations, enabling us to
reduce the time complexity of t-SNE to $\mathcal{O}(N)$. This allows modern
t-SNE implementations to be leveraged for the visulization of data sets
containing up to millions of data points.

@@

\subsubsection*{Repulsive Forces}
The repulsive term can similarily be approximated, motivated by methods from
particle simulations. \citet{van2014accelerating} proposed an approach based on
N-body simulations and used a space-partitioning Barnes-Hut tree approach to
approximate the interaction between data points. This reduces the time
complexity from $\mathcal{O}(N^2)$ to $\mathcal{O}(N \log N)$. More recently,
\citet{linderman2019fast} proposed an alternative approach, FIt-SNE, based on
non-uniform convolutions which further reduces the time complexity to
$\mathcal{O}(N)$.

@@

Similarly to standard t-SNE, a direct calculation of gradients takes
$\mathcal{O}(N \cdot M)$ time, but it is straightforward to adapt the Barnes-Hut
and FIt-SNE approximation schemes, reducing the time complexity to
$\mathcal{O}(M \log N)$ and $\mathcal{O}(M)$, respectively. Special care must be
taken to tune the learning rate during optimization as the existing values
$\eta=200$ and $\eta=N/12$ result in unstable optimization.

@@

\citet{kobak2019umap} showed that the initialization of force-directed layout
algorithms, such as t-SNE, UMAP~\citep{2018arXivUMAP}, and ForceAtlas2~\citep{jacomy2014forceatlas2}, largely dictates the global
consistency of resulting embeddings. When initialized randomly, clusters are
often arbitrarily scattered in the embedding space. However, when using
initalization schemes based on PCA or spectral embeddings, the clusters
identified in the resulting embedding are typically grouped in a globally
coherent manner.

@@

Embeddings produced by standard t-SNE often makes use all available space and
separates clusters by only thin boundaries. When working with large data sets,
this often obscures the global relationships between clusters as all neighboring
groups appear equidistant from one another.  Other dimensionality reduction
methods, such as UMAP or
ForceAtlas2, produce embeddings where clusters
appear more compact and the white-space separating the clusters may be
interpreted as a loose measure of distance. Recently, \citet{bohm2020unifying}
showed that the exaggeration factor $\rho$ could be used to produce layouts more
similar to UMAP and ForceAtlas2. By incorporating exaggeration into later phases
of the optimization, t-SNE introduces more white-space between clusters which better
reflects the global relations between clusters.

@@

\opentsne\ is compatible with the Python data science ecosystem (e.g.
\pkg{numpy}, \pkg{scikit-learn}, \pkg{scanpy}), providing a familiar
and intuitive API to users. Its modular design encourages extensibility and
experimentation with various parameter settings and changes to the analysis
pipeline. \opentsne\ aims to make the latest theoretical advances in t-SNE
accessible to the wider data-science community. These features are easily
accessible through an intuitive API, which closely follows the style
\pkg{scikit-learn}~\citep{sklearn_api}, a popular and widely used toolkit for
machine learning. The following snippet exemplifies this interface and
demonstrates how \opentsne\ can be used to quickly generate several
visualizations, each highlighting different aspects of the underlying data
structure.
\begin{CodeChunk}
\begin{CodeInput}
from openTSNE import TSNE

@@

We illustrate this point by generating four different embeddings of the data on
single-cell gene expression in mouse brain~\citep{tasic2018shared}.
Fig.~\ref{fig:tasic}.a shows an embedding using default t-SNE parameters. While
different clusters of excitatory and inhibitory neurons appear close to one
another, all clusters appear equidistant from their neighbors, and the overall
relations between groups are not obvious. The embedding in Fig.~\ref{fig:tasic}.b
focuses on preserving larger neighborhoods of points, resulting in a more
globally consistent layout where relations between clusters become more
apparent. Here, it is evident from the increased white space between groups that
there is one large class of excitatory neurons and two related classes of
inhibitory neurons. Unfortunately, focusing on preserving large neighborhoods
leads to the absorption of smaller clusters into larger ones. Alternatively,
Fig.~\ref{fig:tasic}.c uses multi-scale similarity kernels that aim to preserve
both the global organization of clusters and prevent smaller cluster absorption. We
constructed the embedding from Fig.~\ref{fig:tasic}.d with the settings used
for Fig.~\ref{fig:tasic}.a, but at a finer level of resolution.
The figure demonstrates that some clusters are composed of numerous, smaller
subgroups representing different cell subpopulations which are not visible under
standard parameter settings.

@@

The results in this paper were obtained using \proglang{Python}~3.7.6 with the
\pkg{openTSNE}~XXX package. \opentsne\ has three direct dependencies:
\pkg{numpy}, \pkg{scipy}, and \pkg{scikit-learn}. In our experiments, we used
\pkg{numpy}~v1.18.1, \pkg{scipy}~v1.4.1. \pkg{scikit-learn}~v0.23.1. As these
packages include their own downstream dependencies, the full environment used to
produce the figures is included in the accompanying GitHub repository
\url{https://github.com/pavlin-policar/opentsne-paper}.

>>>>

The ever-growing volumes of high-dimensional data sets in machine learning calls
for efficient dimensionality reduction techniques and their implementations,
capable of producing informative data visualizations. Popular approaches include
principal component analysis, multidimensional scaling, $t$-distributed stochastic
neighbor embedding (t-SNE)~\citep{maaten2008visualizing}, and uniform manifold
approximation and projections (UMAP)~\citep{2018arXivUMAP}. Among these, t-SNE
has received much attention as it can address high volumes of data and reveal
the underlying data structure. For instance, t-SNE is widely used in the
bioinformatics community in areas such as single-cell
transcriptomics~\citep{macosko2015highly,cao2019single,tasic2018shared}, human
genetics~\citep{hirata2019genetic}, metagenomic
assembly~\citep{beaulaurier2018metagenomic}, the spatial organization of
microbial communities~\citep{sheth2019spatial}, and
metabolomics~\citep{tkachev2019differences}. Reports on single-cell gene
expression data, our running example, often start with an overview of the cell
landscape, where t-SNE is used to embed high-dimensional expression profiles
into a two-dimensional space. Figs.~\ref{fig:macosko}.a and \ref{fig:macosko}.b
show two such embeddings.

@@

Despite its utility, t-SNE has often been criticized for its limited
scalability, lack of global organization -- t-SNE identifies well-defined
clusters that may be arbitrarily scattered throughout the low-dimensional space
-- and the absence of theoretically-founded methods to map new data into
existing embeddings~\citep{ding2018interpretable,becht2019dimensionality}. Most
of these shortcomings have recently been addressed. \citet{linderman2019fast}
developed FIt-SNE, an efficient approximation scheme that massively improves
the scalability of t-SNE, achieving linear time complexity in the number of
samples. \citet{kobak2019art} proposed several techniques to improve global
cluster coherence, including estimating similarities with a mixture of Gaussian
kernels. In our previous work, we introduced a principled approach for embedding
new samples into existing visualizations~\citep{policar2019embedding}.

@@

\subsection{$t$-distributed stochastic neighbor embedding} \label{sec:meth.tsne}

$t$-distributed stochastic neighbor embedding (t-SNE) is a non-linear
dimensionality reduction method that to aims to find a low-dimensional embedding
where neighborhoods are preserved. More formally, given a multi-dimensional data set
$\mathbf{X} = \left \{ \mathbf{x}_1, \mathbf{x}_2, \dots, \mathbf{x}_N \right \}
\in \mathbb{R}^D$ where $N$ is the number of data points in the data set, t-SNE
aims to find a low dimensional embedding $\mathbf{Y} = \left \{ \mathbf{y}_1,
\mathbf{y}_2, \dots, \mathbf{y}_N \right\} \in \mathbb{R}^d$ where $d \ll D$,
such that if points $\mathbf{x}_i$ and $\mathbf{x}_j$ are close in the
high-dimensional space, their corresponding embeddings $\mathbf{y}_i$ and
$\mathbf{y}_j$ are also close. Since t-SNE is primarily used as a visualization
tool, $d$ is typically set to two. The similarity between two data points in the
high-dimensional space is defined using the Gaussian kernel
\begin{equation}
p_{j \mid i} = \frac{\exp \left ( -\frac{1}{2} \mathcal{D}(\mathbf{x}_i, \mathbf{x}_j ) / \sigma_i^2 \right )}
{\sum_{k \neq i } \exp \left ( -\frac{1}{2} \mathcal{D}(\mathbf{x}_i, \mathbf{x}_k ) / \sigma_i^2 \right )}, \quad p_{i \mid i} = 0 \notag
\label{eq:gaussian_kernel}
\end{equation}
where $\mathcal{D}$ is some distance measure. The values $p_{j \mid i}$
are then symmetrized to
\begin{equation}
p_{ij} = \frac{p_{j \mid i} + p_{i \mid j}}{2N}.
\label{eq:symmetrize}
\end{equation}

The bandwidth $\sigma_i$ of each Gaussian kernel is selected such that the perplexity of the distribution matches a user-specified parameter value
\begin{equation}
\text{Perplexity} = 2^{H(P_i)} \notag
\end{equation}
where $H(P_i)$ is the Shannon entropy of $P_i$,
\begin{equation}
H(P_i) = -\sum_i p_{j \mid i} \log_2 (p_{j \mid i}). \notag
\end{equation}
This enables t-SNE to adapt to the varying density of the data in the
multi-dimensional space. The perplexity can be interpreted as the continuous
analogue to the number of nearest neighbors to which the distances will be
preserved.

@@

Optimization is performed using batch gradient descent using the delta-bar-delta
update rule~\citep{jacobs1988increased}. The t-SNE optimization procedure
consists of two phases: in the first \textit{early exaggeration} phase, the
attractive forces between data points are increased by some factor $\rho$,
typically set to $12$, so that points in the embedding can more easily move
throughout the space and settle near their respective neighbors. In the second
phase of the optimization, the attractive forces are then reverted to their
original values with $\rho=1$.

@@

A direct evaluation of t-SNE gradients requires $\mathcal{O}(N^2)$ operations,
which makes its application impractical to any reasonably-sized data set and
highlights the need for the development of efficient approximation schemes.
\citet{van2014accelerating} observed that the t-SNE gradient can be cast as an
N-body problem where data points represent particles that attract and repel each
other. The gradient from Eqn.~(\ref{eq:tsne_gradient}) can be rewritten as
\begin{equation}
\frac{\partial C}{\partial \mathbf{y}_i} = 4 \Bigg [
\underbrace{\sum_{j \neq i} p_{ij} q_{ij} Z \left ( \mathbf{y}_i - \mathbf{y}_j \right )}_{\text{attractive forces}}  -
\underbrace{\sum_{j \neq i} q_{ij}^2 Z \left ( \mathbf{y}_i - \mathbf{y}_j \right )}_{\text{repulsive forces}}
\Bigg ], \label{eq:grad_attr_rep} \notag
\end{equation}
where $Z = \sum_{k \neq l}\left ( 1 + || \mathbf{y}_k - \mathbf{y}_l ||^2 \right
)^{-1}$. We can view this equation as a particle simulation, where the
two terms represent the attractive and repulsive forces between individual
particles. Each term lends itself to efficient approximations, enabling us to
reduce the time complexity of t-SNE to $\mathcal{O}(N)$. This allows modern
t-SNE implementations to be leveraged for the visualization of data sets
containing up to millions of data points.

@@

\subsubsection*{Repulsive Forces}
Similarly, the repulsive term can also be approximated, motivated by methods from
particle simulations. \citet{van2014accelerating} proposed an approach based on
N-body simulations and used a space-partitioning Barnes-Hut tree approach to
approximate the interaction between data points. This reduces the time
complexity from $\mathcal{O}(N^2)$ to $\mathcal{O}(N \log N)$. More recently,
\citet{linderman2019fast} proposed an alternative approach, FIt-SNE, based on
non-uniform convolutions which further reduces the time complexity to
$\mathcal{O}(N)$.

@@

Similarly to standard t-SNE, a direct calculation of gradients takes
$\mathcal{O}(N \cdot M)$ time, but it is straightforward to adapt the Barnes-Hut
and FIt-SNE approximation schemes, reducing the time complexity to
$\mathcal{O}(M \log N)$ and $\mathcal{O}(M)$, respectively. Special care must be
taken to tune the learning rate during optimization as the default values
$\eta=200$ and $\eta=N/12$ result in highly unstable optimization.

@@

\citet{kobak2019umap} showed that the initialization of force-directed layout
algorithms, such as t-SNE, UMAP~\citep{2018arXivUMAP}, and ForceAtlas2~\citep{jacomy2014forceatlas2}, largely dictates the global
consistency of resulting embeddings. When initialized randomly, clusters are
often arbitrarily scattered in the embedding space. However, when using
initialization schemes based on PCA or spectral embeddings, the clusters
identified in the resulting embedding are typically grouped in a globally
coherent manner.

@@

Embeddings produced by standard t-SNE often make use of all available space and
separates clusters by only thin boundaries. When working with large data sets,
this often obscures the global relationships between clusters as all neighboring
groups appear equidistant from one another.  Other dimensionality reduction
methods, such as UMAP and ForceAtlas2, produce embeddings where clusters appear
more compact and the white space separating the clusters may be interpreted as a
loose measure of distance. Recently, \citet{bohm2020unifying} showed that the
exaggeration factor $\rho$ could be used to produce layouts more similar to UMAP
and ForceAtlas2. By incorporating exaggeration into later phases of the
optimization, t-SNE introduces more white space between clusters which better
reflects the global relations between clusters.

@@

\opentsne\ is compatible with the Python data science ecosystem (e.g.
\pkg{numpy}, \pkg{scikit-learn}, \pkg{scanpy}), providing a familiar
and intuitive API to users. Its modular design encourages extensibility and
experimentation with various parameter settings and changes to the analysis
pipeline. \opentsne\ aims to make the latest theoretical advances in t-SNE
accessible to the wider data science community. These features are easily
accessible through an intuitive API, which closely follows the style
\pkg{scikit-learn}~\citep{sklearn_api}, a popular and widely used toolkit for
machine learning. The following snippet exemplifies this interface and
demonstrates how \opentsne\ can be used to quickly generate several
visualizations, each highlighting different aspects of the underlying data
structure.
\begin{CodeChunk}
\begin{CodeInput}
from openTSNE import TSNE

@@

We illustrate this point by generating four different embeddings of the data on
single-cell gene expression in mouse brain~\citep{tasic2018shared}.
Fig.~\ref{fig:tasic}.a shows an embedding using default t-SNE parameters. While
different clusters of excitatory and inhibitory neurons appear close to one
another, all clusters appear equidistant from their neighbors, and the overall
relations between groups are not obvious. The embedding in Fig.~\ref{fig:tasic}.b
focuses on preserving larger neighborhoods of points, resulting in a more
globally consistent layout where relations between clusters become more
apparent. Here, it is evident from the increased white space between groups that
there is one large class of excitatory neurons and two related classes of
inhibitory neurons. Unfortunately, focusing on preserving large neighborhoods
leads to the absorption of smaller clusters into larger ones. Alternatively,
Fig.~\ref{fig:tasic}.c uses multi-scale similarity kernels that aim to preserve
both the global organization of clusters and prevent smaller cluster absorption. We
constructed the embedding from Fig.~\ref{fig:tasic}.d with the settings used
for Fig.~\ref{fig:tasic}.a, but at a finer level of resolution.
The figure demonstrates that some clusters are composed of numerous, smaller
subgroups representing different cell subpopulations that are not visible under
standard parameter settings.

@@

The results in this paper were obtained using \proglang{Python}~3.7.6 with the
\pkg{openTSNE}~v0.6.0 package. \opentsne\ has three direct dependencies:
\pkg{numpy}, \pkg{scipy}, and \pkg{scikit-learn}. In our experiments, we used
\pkg{numpy}~v1.18.1, \pkg{scipy}~v1.4.1. \pkg{scikit-learn}~v0.23.1. As these
packages include their own downstream dependencies, the full environment used to
produce the figures is included in the accompanying GitHub repository
\url{https://github.com/pavlin-policar/opentsne-paper}.
